Cross-validation of regression models: mixed model examples (edited v2)

### 2023-10-15 scratch/gm_addition_to_cv-edited_v2.Rmd
Further edits to mixed model portion of vignette

```{r packages, include=FALSE}
library(cv)
library(car)
library(lattice)
library(latticeExtra)
library(lme4)
```
The following artificial data set exemplifies a difference between cluster-based
and individual-based cross-validation in hierarchical models.

It also shows how, in contrast with ordinary least-squares
models, the variance at different levels of a hierarchical
model need not necessarily become smaller as the model becomes larger.

In this example, the simplest random intercept model results in a between-cluster variance
that is much smaller than that of a model that also includes $x$ as a predictor.
Consequently, different methods of assessing model fit may result in
different best models, underscoring the importance of assessing hierarchical models
at different levels.

Adjusting the parameters generating this data can be used to create examples
that illustrates problems of non-convergence with hierarchical models.
Fitted models with relatively small between-cluster variance, even those using simple
random intercept models, are prone to non-convergence due to the
nature of optimizing algorithms used in `lme` and `lmer`.

@@

At present, the results below are based on a data set that does not
produce non-convergence and are limited to comparisons obtained
by varying three factors:

- Three models: random intercept only, adding x as a predictor, and also adding the group means of x as a predictor,
- The method used for fitting: `lmer` in `lme4` or `lme` in `nlme`,
- Cross-validation resampling individual cases or resampling clusters.

@@

With this data set, the two functions `lme` and `lmer` converged for all models and produced consistent results.
The differences between these two functions, and differences that depend on the optimizer used
within each function, would manifest themselves if the data were generated with a smaller
between-cluster variance.

We visualize predicted values using population predictions with the fixed effects model only
and using BLUPs for random effects.  Since `lme` and `lmer` produced almost identical results, the
figures below are produced only with the `lmer` results.

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(fit0, re.form = ~ 0)
  fit_mod0.re <- predict(fit0)
  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
  fit_mod1.re <- predict(fit1)
  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
  fit_mod2.re <- predict(fit2)
  fit_mod3.fe <- predict(fit3, re.form = ~ 0)
  fit_mod3.re <- predict(fit3)
})

@@

We carry out cross-validation resampling individuals within clusters, based on
fixed and predicted random effects, and resampling clusters, based on
fixed effects only.
In order to reduce between-model variability in comparisons of models,
the `cv` method applied to the list of models created by the `models` function,
performs cross-validation with the same training and validation sets
for each model in order to reduce the between-model variability in the cross-validation
criterion that stems from the random variation in training set selection.

@@

With this example, `lme` and `lmer` give similar results, but resampling cases versus resampling clusters provides
very different comparisons of models.  The model with X alone, without the contextual mean of X, is assessed as
fitting very poorly sampling clusters and, relatively, much better resampling individual cases.  This illustrates the
need to perform cross-validation at different levels.

>>>>

```{r packages, include=FALSE}
library(cv)
library(car)
library(lattice)
library(latticeExtra)
library(lme4)
```
The following artificial data set exemplifies striking aspects of cross-validation
with hierarchical models. Model comparisons using cluster-based versus individual-based 
cross-validation may not agree on a best model. Furthermore, commonly-used measures
of fit, such as least-squares, do not necessarily become smaller as 
models becomes larger.

Varying the parameters of the model, particularly the between-subject variance
relative to the within-subject variance, can also reveal strong differences in the
behaviour of different algorithms used to fit mixed models. These
issues are not pursued in any depth in this vignette.

This example exhibits both phenomena described above. Consider a researcher studying the improvement
in a skill, yodelling for example, among students enrolled in a four-year
yodelling program. The plan is to measure each student's skill level
at the beginning of the program and every year thereafter until
the end of the program, resulting in five annual measurements for each student.

It turns out that yodelling appeals to students of all ages.
Students enrolling in the program range in age from 20 to 70.
Moreover, participants' untrained yodelling skill is similar at all ages, as is
their rate of progress with training. All students complete the
five-year program.

The researcher, with more expertise in yodelling than modelling,
decides to model the response $Y$, yodelling skill, against
age, $X$, reasoning that students get older during their stay in the program,
and age can serve as a proxy for elapsed time.
The researcher knows that a mixed model should be used to account for clustering
due to the expected similarity of measurements taken from each student.

The figures below show the pattern of responses expected from a representative
subset of individuals.

The between-student effect of age is zero but the within-student effect is 1.
Due to the large variation in ages between students, a least-squares model
fitting age as a predictor would produce an estimated slope that is close to
zero, since the estimated slope is heavily weighted toward the between-student effect.
Consider a mixed model using the same fixed-effects model with the addition of a random intercept.

Depending on the estimated variance parameters of the model, a mixed model will apply 
varying degrees of
shrinkage to the random intercept BLUPs [insert reference to ARA&GLM] that
correspond to the variation in the heights of the parallel fitted lines for each student, measured
at a common value of age.

In the example below, the model applies little shrinkage, allowing a large variability in the heights
of fitted lines, which approach closely the observed values for each student.
Consequently, the model's fit is similar to that of
a fixed-effects model with an effect of age together with a categorical predictor for 
individual students. Thus, the model fits individual observations well, resulting in a
favourable assessment using individual-based cross-validation. In contrast, the very large
variability in the BLUPs results in large residuals for
cluster-based cross-validation, producing an
unfavourable assessment compared with the smaller model using only a random intercept.

By comparison, if the mixed model had applied considerable shrinkage, neither the
between-cluster nor the between-individual cross-validations would have shown much 
improvement over the random-intercept only model.  In our experience, the degree of
shrinkage does not vary smoothly as parameters are changed but tends to be
"all or nothing." Its behaviour is affected considerably by the choice of the
algorithm used to fit the model.

Finally, a model that also includes a compositional variable consisting of the mean age
of each student, results in the random intercept BLUPs representing variation in the students' fitted
lines measured at the mean age of each student. Thus, with this example, both 
the cluster-based and the individual-based fits are relatively good.

To serve as a comparison, we also include cross-validation using the model known to 
have generated the data, namely a model with a between-subject age effect equal to zero by
modelling the response $Y$ with a fixed effects model using an age centered within
each student, i.e. the age at assessment less the mean age of each student's 
assessments, plus a random term for each student.

A closely related question is that of non-convergence of mixed-model
algorithms. Adjusting the parameters generating this data can be used to create examples
illustrating problems of non-convergence with hierarchical models.
Fitted models with relatively small between-cluster variance, even those using simple
random intercept models, are prone to non-convergence due to the
nature of optimizing algorithms used in `lme` and `lmer`.

@@

At present, the results below are based on a data set that does not
produce non-convergence and are limited to comparisons obtained
by varying two factors:

- Four models: random intercept only, adding x as a predictor, adding the 
  group means of x (xm) as a predictor, and
  a fourth model with a random intercept and x centered within each student.
- Cross-validation resampling individual cases or resampling clusters.

@@

We visualize predicted values using population predictions with the fixed effects model only
and using BLUPs for random effects. 
```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(fit0, re.form = ~ 0)
  fit_mod0.re <- predict(fit0)
  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
  fit_mod1.re <- predict(fit1)
  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
  fit_mod2.re <- predict(fit2)
  fit_mod3.fe <- predict(fit3, re.form = ~ 0)
  fit_mod3.re <- predict(fit3)
})

@@

We carry out cross-validation resampling individuals within clusters, based on
fixed and predicted random effects, and resampling clusters, based on
fixed effects only.
In order to reduce between-model variability in comparisons of models,
the `cv` method applied to the list of models created by the `models` function,
performs cross-validation with the same training and validation sets
for each model.

@@

In summary, the model with $X$ alone, without the compositional mean of $X$, is assessed as
fitting very poorly sampling clusters but, relatively, much better resampling individual cases.
The model that includes both $X$ and its compositional mean produces better results using both
criteria: resampling individual cases and resampling cluster, illustrating the need to
assess models at different levels of resampling.
